Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2022-05-16 chapters/Paper_one.tex

\chapter{Lattice Boltzmann method for thin-liquid-film hydrodynamics}
\label{chapter:first_paper}

@@

\section{Numerical model}\label{sec:method}

When a layer of fluid is characterized by a vertical length scale $H$ much
smaller than the longitudinal one $L$, the equations of motion can be
simplified under the approximation that the ratio of the length scales, 
$\varepsilon \equiv H/L$, is small ($\varepsilon \ll1$, see Fig.~\ref{fig:scheme}).
\begin{figure}
    \centering
    \includegraphics[width = 0.45\textwidth]{graphics/Fig_1_Scheme_V3_filters_2.png}
    \caption{(Color online) Schematic sketch of a model system: a thin liquid film deposited on a flat substrate. The air-liquid interface is represented by the {\it height} $h(x,y,t)$. The characteristic thickness of the film is given by $H$.}
    \label{fig:scheme}
\end{figure}
In this limit, and for small reduced Reynolds number, $\varepsilon^2Re$
(where $Re=\frac{U L}{\nu}$, with $U$ being a characteristic velocity of the fluid system and $\nu$ being the fluid's kinematic viscosity),
the lubrication approximation tells that the dynamics is governed by equation (\ref{eq:lubr}).
Instead of directly solving Eq.~(\ref{eq:lubr}) numerically, we follow an alternative strategy.
We build our numerical model on a class of LBMs originally proposed as solvers for
the shallow water
equations~\cite{Salmon:1999:0022-2402:503,PhysRevE.65.036309,zhou2004lattice,van2010study}.
The lattice Boltzmann equation for the discrete probability density functions of a fluid system
subject to a total force (that can include both internal and external forces)
$\mathbf{F}_{\mbox{\tiny{tot}}}$,
$f_l(\mathbf{x},t)$, reads:
\begin{equation}\label{eq:LBE}
\begin{split}
&f_l(\mathbf{x}+\mathbf{c}^{(l)}\Delta t,t+\Delta t) = \\
&(1 - \omega) f_l(\mathbf{x},t) + \omega f_l^{(eq)}(\mathbf{x},t) + w_l \frac{\Delta t}{c_s^2} \mathbf{c}^{(l)} \cdot \mathbf{F}_{\mbox{\tiny{tot}}},
\end{split}
\end{equation}
where $l$ labels the lattice velocities $\mathbf{c}_l$ and runs from $0$ to $Q-1$, with $Q$ being the number
of velocities characterizing the scheme.
Algorithmically, this equation can be seen as made up of two steps. A {\it local collision}
step where the $f_l(\mathbf{x},t)$ ``relax'' towards the local equilibrium distributions
$f^{(eq)}_l(\mathbf{x},t)$ with rate  
$\omega = \Delta t/\tau$ (where $\tau$, the relaxation time, is proportional to the kinematic viscosity $\nu$): the distribution functions are substituted by their weighted average 
(with weights $\omega$ and $1-\omega$) 
with the equilibria, with an added so-called "source" term (the last term on the right hand side of Eq. (\ref{eq:LBE})), when a force is present. A {\it non-local streaming} step where the updated distribution functions are scattered to the nearest neighbouring sites. 
%The last term on the rhs of (\ref{eq:LBE}) includes the total force $\mathbf{F}_{\mbox{\tiny{tot}}}$ which can be in general the sum of bulk and internal (possibly space and time dependent) contributions. 
The parameters $c_s$ (the lattice speed of sound) and $w_l$ (the 
so called "weights")
depend on the geometry of the lattice and are determined under suitable constraints on 
the form of the tensorial moments in the lattice velocities up to fourth order~\cite{wolf-gladrow}.
We work with two-dimensional square lattices of side length $N\Delta x$, with lattice constant $\Delta x$ and $Q=9$.
For simplicity, we keep $\Delta t=\Delta x=1$ throughout this paper and follow the standard notation, where $c_s = 1/\sqrt{3}$ and the $\mathbf{c}^{(l)} = (c^{(l)}_x,c^{(l)}_y)$, $l=0,1,\dots,8$, are~\cite{qian-dhumieres-lallemand,shan_yuan_chen_2006}

@@

With the usual {\it ansatz}

@@

\Pi(h) = \kappa f(h) &=& \underbrace{(1 - \cos(\theta))\frac{(n-1)(m-1)}{(n-m)h_*}}_{\kappa}\times
\nonumber\\&&\underbrace{\left[\left(\frac{h_*}{h}\right)^n -\left(\frac{h_*}{h}\right)^m\right]}_{f(h)}.

@@

Before concluding this section, let us notice that special care has to be taken
in the implementation of the numerical scheme, when evaluating the
forcing term since it contains higher order derivatives (the gradient $p_{\mbox{\tiny{film}}}$,
which in turn includes the Laplace
 pressure $\gamma \nabla^2 h$, see Eq.(~\ref{eq:pfilm}) and, hence,
 spurious lattice effects may arise.  
We noticed, for example, that a centered scheme to calculate gradients \cite{zhou2004lattice}
does not guarantee the sufficient degree of isotropy on the lattice as, e.g., for the relaxation of
a droplet (discussed in section~\ref{sec:results}), where it led to unphysical droplet shapes.
Therefore, we use the following expressions to compute the gradients
\begin{equation}\label{eq:derivative}
    \nabla \phi(\mathbf{x}) = 3\sum_{l=0}^8w_l\mathbf{c}^{(l)} \phi(\mathbf{x}+\mathbf{c}^{(l)})+O(\nabla^3),
\end{equation}
and the Laplacian 
\begin{align}\label{eq:n_laplace}
    \nabla^2 \phi(\mathbf{x}) =& \frac{1}{6}\bigg[4\sum_{l = {\rm odd}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)})\nonumber\\ 
    &+ 1\sum_{l = {\rm even}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)}) - 20\phi(\mathbf{x}) \bigg] + O(\nabla^4),
\end{align}
respectively \cite{doi:10.1137/S1064827599357188,THAMPI20131}, for a generic scalar field $\phi$ 
(be it the height field $h$, the pressure $p_{\mbox{\tiny{film}}}$ or a position dependent surface tension field).
Besides the higher degree of the isotropy, the scheme 
(\ref{eq:derivative}-\ref{eq:n_laplace}) has the advantage of employing directly the set of lattice Boltzmann speeds.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{graphics/Fig_2_1_rti_renders_a).png}
    \includegraphics[width=0.5\textwidth]{graphics/Fig_2_2_rti_renders_b).png}
    \includegraphics[width=0.5\textwidth]{graphics/Fig_2_3_rti_renders_c).png}
    \includegraphics[width=0.5\textwidth]{graphics/Fig_2_4_rti_renders_d).png}
  \caption{(Color online) Time evolution of the free surface for the Rayleigh-Taylor instability at $\tau_{\mbox{\tiny{cap}}}\approx50, 75, 100, 188$, corresponding to the lattice Boltzmann time steps [9000, 14000, 19000, 35000] $\Delta t$ in (a) - (d) respectively. For a more clear visualization we only show a small patch of size $256\times256$ centered in the middle of the $2048\times 2048$ domain. The fluctuations of earlier states still follow the linear stability analysis (See Fig.~\ref{fig:RTI} for the the power
spectrum  of the  height fluctuations versus wavenumber).}
  \label{fig:RTI_evolution}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_3_new_RTI_spectra_single_color_same_tau_capkc_correct_rescaled_x_axis_inset.png}
    \caption{(Color online) Power spectrum of the height fluctuations versus wavenumber.} The different colors and symbols belong to different values of graviational acceleration, $g=4\cdot 10^{-5}$ is given by blue circles (\textcolor{pyblue}{$\bullet$}), $g=6\cdot 10^{-5}$ by orange triangles (\textcolor{pyorange}{$\blacktriangle$}) and $g=8\cdot 10^{-5}$ is given by green squares (\textcolor{pygreen}{$\blacksquare$}). Same colored lines are taken at different time steps. In the inset we show the growth rate $\sigma(k)$ for the largest value of $g$ (symbols) and the theoretical growth rate according to Eq.~(\ref{eq:RTgrowth}) (solid line).
    \label{fig:RTI}
\end{figure}

@@

\begin{figure}
    \centering %Figure without inset is also available, however I personally like this more.
    \includegraphics[width=0.75\textwidth]{graphics/Fig_5_Indirect_Cox_Voinov_all_data_visually_appealing_slip_2_m_nosci.png}
    \caption{(Color online) Difference of cubed instantaneous and equilibrium contact angles, $\theta_{num}^3-\theta_{eq}^3$, vs. capillary number $Ca$ for a spreading droplet; the dashed line shows a linear dependence (consistent with the Cox-Voinov law). The different symbols represent different viscosities, while the dashed line is a linear function of the capillary number.
    \label{fig:Cox-Voinov}}
\end{figure}
The spreading dynamics can be investigated even more quantitatively in terms of the 
so-called Cox-Voinov law and Tanner's law~\cite{Tanner_1979}. The first one relates the apparent contact angle to 
the velocity $U$ of the spreading front (the contact line), at various times, by 
$\theta^3 - \theta_{eq}^3 \propto Ca$. The capillary number $Ca$ is defined
as $Ca=\mu U/\gamma$~\cite{doi:10.1146/annurev-fluid-011212-140734}.
In Fig.~\ref{fig:Cox-Voinov} we plot $\theta^3(t) - \theta_{eq}^3$ vs $Ca(t)$ from a numerical simulation of a spreading drop: a good linear scaling, in agreement with the Cox-Voinov law, is 
observed, as highlighted by the dashed line.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_6_Tanners_law_slip_2_paper_rescaled_t.png}
    \caption{(Color online) Time evolution of the droplet base radius of a spreading droplet; the dashed red line shows a $\tilde{t}^{1/10}$ power law (In consistence with Tanner's law). As in Fig.~\ref{fig:Cox-Voinov} different symbols refer to different viscosities. The radius clearly grows with the predicted power law until it saturates. Upon rescaling the time with $\tau_{\mbox{\tiny{cap}}}$ the curves of all three viscosities collapse into a single one.}
    \label{fig:Tanners_law}
\end{figure}
The Tanner's law which states that the radius of the droplet grows with time as
\begin{equation}
\label{eq:tanners_law}
    R(t)\approx \left[\frac{10\gamma}{9B\mu}\left(\frac{4V}{\pi}\right)^3 t\right]^{1/10},
\end{equation}
with the constant $B$ being such that 
$B^{1/10} \approx 1.2$.  
In Fig.~\ref{fig:Tanners_law} 
we plot the measured radius of the droplet divided by its initial radius $R_0$ as a function of the dimensionless time $\tilde{t} = t/\tau_{\mbox{\tiny{cap}}}$ 
(here $\tau_{\mbox{\tiny{cap}}} = \frac{\mu R}{\gamma}$). For the three viscosities considered 
in Figs.~\ref{fig:Cox-Voinov},\ref{fig:Tanners_law} our capillary times are 
$\tau_{\mbox{\tiny{cap}}} = [1333, 1667, 2333] \Delta t$. 
We see a saturation at $R/R_0=1.17$ because the droplet is very close to 
its equilibrium shape. During the growth phase the radius follows indeed a power law 
in $\tilde{t}$ with exponent $1/10$, which is shown by the red dashed line, in agreement with Tanner's prediction and 
experimental results \cite{rioboo2002time, jambon-puillet_carrier_shahidzadeh_brutin_eggers_bonn_2018, cazabat1986dynamics, CHEN198860}.
We further notice that within our
simulations the droplet needs about $12\tau_{\mbox{\tiny{cap}}}$ to reach its equilibrium shape.

\subsection{A Sliding droplet}
As a further validation case we consider the sliding of a droplet on an inclined plane.
For a droplet to slide over an inclined plane, a minimum tilting angle $\alpha >0$ of the substrate is required~\cite{Furmidge}, which in our case is due to the friction term Eq.~(\ref{eq:alphafric}). Until this critical angle is reached energy is stored in the deformation of the surface as the upper left inset in Fig.~\ref{fig:CaBo} shows. Above such a critical angle, a linear relation between the terminal sliding velocity $U_{\infty}$ and the gravitational force $\propto m g \sin \alpha$ is observed~\cite{Podgorski,Kim,Sbragaglia}; in dimensionless numbers such 
behaviour is expressed by 
\begin{equation}\label{eq:CaBo}
Ca \propto Bo - Bo_c,
\end{equation}
where the capillary number is based on $U_{\infty}$ and $Bo$ is the so called Bond number, given by
\begin{equation}
  Bo = (3V/4\pi)^{2/3}\rho g \frac{\sin\alpha}{\gamma}.
\end{equation}
$Bo_c$ is the critical Bond number, defined in terms of the critical 
tilting angle $\alpha_c$.
  \begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_7_Ca_Bo_true_with_pic.png}
    \caption{(Color online) $Ca$ vs $Bo$ for a sliding droplet: notice that a finite minimum forcing (corresponding to $Bo_c$) is needed to actuate the droplet. For $Bo > Bo_c$ a linear relation, $Ca \sim Bo$, is observed. In the insets we show the shape of the droplet for both, the pinned (upper left) as well as the sliding (lower right) case.
    }
    \label{fig:CaBo}
\end{figure}
In Fig.~\ref{fig:CaBo} we plot $Ca$ vs $Bo$ from our numerical simulations, showing that the phenomenology described by Eq.~(\ref{eq:CaBo}) is 
indeed reproduced, i.e. the onset of sliding takes place at a finite 
forcing, beyond which the linear scaling $Ca \sim Bo$ is fulfilled.

\begin{figure}
    \centering
    \includegraphics[width=0.32\textwidth]{graphics/Fig_8_1_logo_v2_a).png}
    \includegraphics[width=0.32\textwidth]{graphics/Fig_8_2_logo_v2_b).png}
    \includegraphics[width=0.32\textwidth]{graphics/Fig_8_3_logo_v2_c).png}
  \caption{Time evolution of the free surface on a chemically patterned substrate on a 512x512 $\Delta x^2$ domain. Varying the contact angle between the letters and the rest of the substrate yields the shown dewetting pattern. The letters are more wettable then the rest. To emphasis the process we use a color gradient raging from dark blue to light blue.
  Starting from a randomly perturbed film height, the fluid starts to dewet the pattern (a) and after 2400$\Delta t$ the letters and a surrounding rim structure are clearly visible. In (b) at 16800$\Delta t$ a depletion around the letters is already very prominent. Towards the end of the simulation (c) at 97200$\Delta t$, the instability of the thin film also leads to film rupture. Holes form between the letters E, R and N.  
  }
  \label{fig:Logo_evolution}
\end{figure}

@@

\section{Computational aspects}
We use \textit{OpenACC} directives to allow our code to run on accelerator devices, such as 
Graphics Processing Units (GPUs), while being able, at the same time, to exploit the well known good scaling properties of the LBM on parallel machines~\cite{Chandrasekaran:2017:OPC:3175812}. 
\textit{OpenACC} is particularly versatile in terms of programmability since it only requires a few lines of code to allow us to harness the power of state of the art accelerators. In this sense \textit{OpenACC} is very similar to \textit{OpenMP} and more readable as well as much easier to program than \textit{CUDA}.

@@

\begin{table}
 \begin{tabular}{|c|c c c c c c |}
 \hline
  Lattice/Accelerator & $128^2$ & $256^2$ & $512^2$ & $1024^2$ & $2048^2$ & $4096^2$ \\ \hline
  GTX 1080TI & $157.6$ & $279.2$ & $382.6$ & $414.9$ & $404.7$ & $395.6$ \\ \hline
  Quadro K2200 & $33.5$ & $42.9$ & $46.6$ & $48.2$ & $49.0$ & $X$ \\ \hline
  i7-4790 & $6.4$ & $5.8$ & $4.5$ & $4.6$ & $4.5$ & $4.3$ \\
  \hline
 \end{tabular}
 \caption{Performance analysis based on a MLUPS measurement. The different columns relate to different lattice sizes, while the rows correspond to the two GPUs and one CPU used. All simulations are run for 100000$\Delta t$ with FP64 double precision.}
 \label{tab:efficiency}
 \end{table}

@@

\begin{acknowledgements}
The authors acknowledge financial support by the Deutsche 
Forschungsgemeinschaft (DFG) within the Cluster of Excellence ``Engineering of Advanced Materials'' (project EXC 315) (Bridge Funding). The work has been partly performed under the Project HPC-EUROPA3 (INFRAIA-2016-1-730897), with the support of the EC Research Innovation Action under the H2020 Programme; in particular, S. Z. gratefully acknowledges the support of Consiglio Nazionale delle Ricerche (CNR) and the computer resources and technical support provided by CINECA.
\end{acknowledgements}

>>>>

\chapter{Lattice Boltzmann method for thin-liquid-film hydrodynamics}
\label{chapter:first_paper}
\epigraph{\textit{Das Schicksal hat mich angelacht und mir ein Geschenk gemacht.}}{RAMMSTEIN}

\textit{\small{This chapter has been published and can be found as: Zitz, S., Scagliarini, A., Maddu, S., Darhuber, A. A., \& Harting, J. (2019). Lattice Boltzmann method for thin-liquid-film hydrodynamics. Physical Review E, 100(3), 033313}}
\\
\\

We propose a novel approach to the numerical simulation of thin film flows,
based on the lattice Boltzmann method. We outline the basic features of the method, show in which
limits the expected thin film equations are recovered and perform validation tests.
The numerical scheme is applied to the viscous Rayleigh-Taylor instability of a
thin film and to the spreading of a sessile drop towards its equilibrium
contact angle configuration. We show that the Cox-Voinov law is satisfied, and
that the effect of a tunable slip length on the substrate is correctly
captured. We address, then, the problem of a droplet sliding on an inclined
plane, finding that the Capillary number scales linearly with the Bond number,
in agreement with experimental results. At last, we demonstrate the ability of
the method to handle heterogenous and complex systems by showcasing the
controlled dewetting of a thin film on a chemically structured substrate.

@@

\section{Numerical Model}\label{sec:method}

When a layer of fluid is characterized by a vertical length scale $H$ much
smaller than the longitudinal one $L$, the equations of motion can be
simplified under the approximation that the ratio of the length scales, 
$\varepsilon \equiv H/L$, is small ($\varepsilon \ll1$, see Fig.~\ref{fig:scheme}).
\begin{figure}
    \centering
    \includegraphics[width = 0.75\textwidth]{graphics/Fig_1_Scheme_V3_filters_2.png}
    \caption{Schematic sketch of a model system: a thin liquid film deposited on a flat substrate. The air-liquid interface is represented by the \textit{height} $h(x,y,t)$. The characteristic thickness of the film is given by $H$.}
    \label{fig:scheme}
\end{figure}
In this limit, and for small reduced Reynolds number, $\varepsilon^2Re$
the lubrication approximation tells that the dynamics is governed by equation (\ref{eq:lubr}).
Instead of directly solving Eq.~(\ref{eq:lubr}) numerically, we follow an alternative strategy.
We build our numerical model on a class of LBMs originally proposed as solvers for
the shallow water
equations~\cite{Salmon:1999:0022-2402:503,PhysRevE.65.036309,zhou2004lattice,van2010study}.
The lattice Boltzmann equation for the discrete probability density functions of a fluid system
subject to a total force (that can include both internal and external forces)
$\mathbf{F}_{\mbox{\tiny{tot}}}$,
$f_l(\mathbf{x},t)$, reads:
\begin{equation}\label{eq:LBE}
\begin{split}
&f_l(\mathbf{x}+\mathbf{c}^{(l)}\Delta t,t+\Delta t) = \\
&(1 - \omega) f_l(\mathbf{x},t) + \omega f_l^{(eq)}(\mathbf{x},t) + w_l \frac{\Delta t}{c_s^2} \mathbf{c}^{(l)} \cdot \mathbf{F}_{\mbox{\tiny{tot}}},
\end{split}
\end{equation}
where $l$ labels the lattice velocities $\mathbf{c}_l$ and runs from $0$ to $Q-1$, with $Q$ being the number
of velocities characterizing the scheme.
Algorithmically, this equation can be seen as made up of two steps. A \textit{local collision}
step where the $f_l(\mathbf{x},t)$ ``relax'' towards the local equilibrium distributions
$f^{(eq)}_l(\mathbf{x},t)$ with rate  
$\omega = \Delta t/\tau$ (where $\tau$, the relaxation time, is proportional to the kinematic viscosity $\nu$): the distribution functions are substituted by their weighted average 
(with weights $\omega$ and $1-\omega$) 
with the equilibria, with an added so-called "source" term (the last term on the right hand side of Eq. (\ref{eq:LBE})), when a force is present. A \textit{non-local streaming} step where the updated distribution functions are scattered to the nearest neighbouring sites. 
The parameters $c_s$ (the lattice speed of sound) and $w_l$ (the 
so called "weights")
depend on the geometry of the lattice and are determined under suitable constraints on 
the form of the tensorial moments in the lattice velocities up to fourth order~\cite{wolf-gladrow}.
We work with two-dimensional square lattices of side length $N\Delta x$, with lattice constant $\Delta x$ and $Q=9$.
For simplicity, we keep $\Delta t=\Delta x=1$ throughout this paper and follow the standard notation, where $c_s = 1/\sqrt{3}$ and the $\mathbf{c}^{(l)} = (c^{(l)}_x,c^{(l)}_y)$, $l=0,1,\dots,8$, are~\cite{qian-dhumieres-lallemand,shan_yuan_chen_2006}

@@

With the usual \textit{ansatz}

@@

\Pi(h) = \kappa f(h) = \underbrace{(1 - \cos(\theta))\frac{(n-1)(m-1)}{(n-m)h_*}}_{\kappa}\underbrace{\left[\left(\frac{h_*}{h}\right)^n -\left(\frac{h_*}{h}\right)^m\right]}_{f(h)}.

@@

Before concluding this section, let us notice that special care has to be taken
in the implementation of the numerical scheme, when evaluating the
forcing term since it contains higher order derivatives (the gradient $p_{\mbox{\tiny{film}}}$,
which in turn includes the Laplace
 pressure $\gamma \nabla^2 h$, see Eq.(~\ref{eq:pfilm}) and, hence,
 spurious lattice effects may arise.  
We noticed, for example, that a centered scheme to calculate gradients \cite{zhou2004lattice}
does not guarantee the sufficient degree of isotropy on the lattice as, e.g., for the relaxation of
a droplet (discussed in section~\ref{sec:results}), where it led to unphysical droplet shapes.
Therefore, we use the following expressions to compute the gradients
\begin{equation}\label{eq:derivative}
    \nabla \phi(\mathbf{x}) = 3\sum_{l=0}^8w_l\mathbf{c}^{(l)} \phi(\mathbf{x}+\mathbf{c}^{(l)})+O(\nabla^3),
\end{equation}
and the Laplacian 
\begin{align}\label{eq:n_laplace}
    \nabla^2 \phi(\mathbf{x}) =& \frac{1}{6}\bigg[4\sum_{l = \text{odd}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)})\nonumber\\ 
    &+ 1\sum_{l = \text{even}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)}) - 20\phi(\mathbf{x}) \bigg] + O(\nabla^4),
\end{align}
respectively \cite{doi:10.1137/S1064827599357188,THAMPI20131}, for a generic scalar field $\phi$ 
(be it the height field $h$, the pressure $p_{\mbox{\tiny{film}}}$ or a position dependent surface tension field).
Besides the higher degree of the isotropy, the scheme 
(\ref{eq:derivative}-\ref{eq:n_laplace}) has the advantage of employing directly the set of lattice Boltzmann speeds.

@@

\begin{figure}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_1_rti_renders_a).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_2_rti_renders_b).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_3_rti_renders_c).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_4_rti_renders_d).png}
  \caption{Time evolution of the free surface for the Rayleigh-Taylor instability at $\tau_{\mbox{\tiny{cap}}}\approx50, 75, 100, 188$, corresponding to the lattice Boltzmann time steps [9000, 14000, 19000, 35000] $\Delta t$ in (a) - (d) respectively. For a more clear visualization we only show a small patch of size $256\times256$ centered in the middle of the $2048\times 2048$ domain. The fluctuations of earlier states still follow the linear stability analysis (See Fig.~\ref{fig:RTI} for the the power
spectrum  of the  height fluctuations versus wavenumber).}
  \label{fig:RTI_evolution}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_3_new_RTI_spectra_single_color_same_tau_capkc_correct_rescaled_x_axis_inset.png}
    \caption{Power spectrum of the height fluctuations versus wavenumber.} The different colors and symbols belong to different values of graviational acceleration, $g=4\cdot 10^{-5}$ is given by blue circles (\textcolor{pyblue}{$\bullet$}), $g=6\cdot 10^{-5}$ by orange triangles (\textcolor{pyorange}{$\blacktriangle$}) and $g=8\cdot 10^{-5}$ is given by green squares (\textcolor{pygreen}{$\blacksquare$}). Same colored lines are taken at different time steps. In the inset we show the growth rate $\sigma(k)$ for the largest value of $g$ (symbols) and the theoretical growth rate according to Eq.~(\ref{eq:RTgrowth}) (solid line).
    \label{fig:RTI}
\end{figure}

@@

\begin{figure}
    \centering 
    \includegraphics[width=0.75\textwidth]{graphics/Fig_5_Indirect_Cox_Voinov_all_data_visually_appealing_slip_2_m_nosci.png}
    \caption{Difference of cubed instantaneous and equilibrium contact angles, $\theta_{num}^3-\theta_{eq}^3$, vs. capillary number $Ca$ for a spreading droplet; the dashed line shows a linear dependence (consistent with the Cox-Voinov law). The different symbols represent different viscosities, while the dashed line is a linear function of the capillary number.
    \label{fig:Cox-Voinov}}
\end{figure}
The spreading dynamics can be investigated even more quantitatively in terms of the 
so-called Cox-Voinov law and Tanner's law~\cite{Tanner_1979}. The first one relates the apparent contact angle to 
the velocity $U$ of the spreading front (the contact line), at various times, by 
$\theta^3 - \theta_{eq}^3 \propto Ca$. The capillary number $Ca$ is defined
as $Ca=\mu U/\gamma$~\cite{doi:10.1146/annurev-fluid-011212-140734}.
In Fig.~\ref{fig:Cox-Voinov} we plot $\theta^3(t) - \theta_{eq}^3$ vs $Ca(t)$ from a numerical simulation of a spreading drop: a good linear scaling, in agreement with the Cox-Voinov law, is 
observed, as highlighted by the dashed line.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_6_Tanners_law_slip_2_paper_rescaled_t.png}
    \caption{Time evolution of the droplet base radius of a spreading droplet; the dashed red line shows a $\tilde{t}^{1/10}$ power law (In consistence with Tanner's law). As in Fig.~\ref{fig:Cox-Voinov} different symbols refer to different viscosities. The radius clearly grows with the predicted power law until it saturates. Upon rescaling the time with $\tau_{\mbox{\tiny{cap}}}$ the curves of all three viscosities collapse into a single one.}
    \label{fig:Tanners_law}
\end{figure}
The Tanner's law which states that the radius of the droplet grows with time as
\begin{equation}
\label{eq:tanners_law}
    R(t)\approx \left[\frac{10\gamma}{9B\mu}\left(\frac{4V}{\pi}\right)^3 t\right]^{1/10},
\end{equation}
with the constant $B$ being such that 
$B^{1/10} \approx 1.2$.  
In Fig.~\ref{fig:Tanners_law} 
we plot the measured radius of the droplet divided by its initial radius $R_0$ as a function of the dimensionless time $\tilde{t} = t/\tau_{\mbox{\tiny{cap}}}$ 
(here $\tau_{\mbox{\tiny{cap}}} = \frac{\mu R}{\gamma}$). For the three viscosities considered 
in Figs.~\ref{fig:Cox-Voinov},\ref{fig:Tanners_law} our capillary times are 
$\tau_{\mbox{\tiny{cap}}} = [1333, 1667, 2333] \Delta t$. 
We see a saturation at $R/R_0=1.17$ because the droplet is very close to 
its equilibrium shape. During the growth phase the radius follows indeed a power law 
in $\tilde{t}$ with exponent $1/10$, which is shown by the red dashed line, in agreement with Tanner's prediction and 
experimental results \cite{rioboo2002time, jambon-puillet_carrier_shahidzadeh_brutin_eggers_bonn_2018, cazabat1986dynamics, CHEN198860}.
We further notice that within our
simulations the droplet needs about $12\tau_{\mbox{\tiny{cap}}}$ to reach its equilibrium shape.

\subsection{A Sliding droplet}
As a further validation case we consider the sliding of a droplet on an inclined plane.
For a droplet to slide over an inclined plane, a minimum tilting angle $\alpha >0$ of the substrate is required~\cite{Furmidge}, which in our case is due to the friction term Eq.~(\ref{eq:alphafric}). Until this critical angle is reached energy is stored in the deformation of the surface as the upper left inset in Fig.~\ref{fig:CaBo} shows. Above such a critical angle, a linear relation between the terminal sliding velocity $U_{\infty}$ and the gravitational force $\propto m g \sin \alpha$ is observed~\cite{Podgorski,Kim,Sbragaglia}; in dimensionless numbers such 
behaviour is expressed by 
\begin{equation}\label{eq:CaBo}
Ca \propto Bo - Bo_c,
\end{equation}
where the capillary number is based on $U_{\infty}$ and $Bo$ is the so called Bond number, given by
\begin{equation}
  Bo = (3V/4\pi)^{2/3}\rho g \frac{\sin\alpha}{\gamma}.
\end{equation}
$Bo_c$ is the critical Bond number, defined in terms of the critical 
tilting angle $\alpha_c$.
  \begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_7_Ca_Bo_true_with_pic.png}
    \caption{$Ca$ vs $Bo$ for a sliding droplet: notice that a finite minimum forcing (corresponding to $Bo_c$) is needed to actuate the droplet. For $Bo > Bo_c$ a linear relation, $Ca \sim Bo$, is observed. In the insets we show the shape of the droplet for both, the pinned (upper left) as well as the sliding (lower right) case.
    }
    \label{fig:CaBo}
\end{figure}
In Fig.~\ref{fig:CaBo} we plot $Ca$ vs $Bo$ from our numerical simulations, showing that the phenomenology described by Eq.~(\ref{eq:CaBo}) is 
indeed reproduced, i.e. the onset of sliding takes place at a finite 
forcing, beyond which the linear scaling $Ca \sim Bo$ is fulfilled.

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{graphics/Fig_8_1_logo_v2_a).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_8_2_logo_v2_b).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_8_3_logo_v2_c).png}
  \caption{Time evolution of the free surface on a chemically patterned substrate on a 512x512 $\Delta x^2$ domain. Varying the contact angle between the letters and the rest of the substrate yields the shown dewetting pattern. The letters are more wettable then the rest. To emphasis the process we use a color gradient raging from dark blue to light blue.
  Starting from a randomly perturbed film height, the fluid starts to dewet the pattern (a) and after 2400$\Delta t$ the letters and a surrounding rim structure are clearly visible. In (b) at 16800$\Delta t$ a depletion around the letters is already very prominent. Towards the end of the simulation (c) at 97200$\Delta t$, the instability of the thin film also leads to film rupture. Holes form between the letters E, R and N.  
  }
  \label{fig:Logo_evolution}
\end{figure}

@@

\section{Computational Aspects}
We use \textit{OpenACC} directives to allow our code to run on accelerator devices, such as 
Graphics Processing Units (GPUs), while being able, at the same time, to exploit the well known good scaling properties of the LBM on parallel machines~\cite{Chandrasekaran:2017:OPC:3175812}. 
\textit{OpenACC} is particularly versatile in terms of programmability since it only requires a few lines of code to allow us to harness the power of state of the art accelerators. In this sense \textit{OpenACC} is very similar to \textit{OpenMP} and more readable as well as much easier to program than \textit{CUDA}.

@@

\begin{table}
 \centering
 \begin{tabular}{|c|c c c c c c |}
 \hline
  Lattice/Accelerator & $128^2$ & $256^2$ & $512^2$ & $1024^2$ & $2048^2$ & $4096^2$ \\ \hline
  GTX 1080TI & $157.6$ & $279.2$ & $382.6$ & $414.9$ & $404.7$ & $395.6$ \\ \hline
  Quadro K2200 & $33.5$ & $42.9$ & $46.6$ & $48.2$ & $49.0$ & $X$ \\ \hline
  i7-4790 & $6.4$ & $5.8$ & $4.5$ & $4.6$ & $4.5$ & $4.3$ \\
  \hline
 \end{tabular}
 \caption{Performance analysis based on a MLUPS measurement. The different columns relate to different lattice sizes, while the rows correspond to the two GPUs and one CPU used. All simulations are run for 100000$\Delta t$ with FP64 double precision.}
 \label{tab:efficiency}
 \end{table}

@@

\section{Acknowledgements}\label{sec:ack}
The authors acknowledge financial support by the Deutsche 
Forschungsgemeinschaft (DFG) within the Cluster of Excellence ``Engineering of Advanced Materials'' (project EXC 315) (Bridge Funding). The work has been partly performed under the Project HPC-EUROPA3 (INFRAIA-2016-1-730897), with the support of the EC Research Innovation Action under the H2020 Programme; in particular, S. Z. gratefully acknowledges the support of Consiglio Nazionale delle Ricerche (CNR) and the computer resources and technical support provided by CINECA.

\section{Appendix}\label{sec:app}
\label{app:only}
\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{graphics/Fig_9_1_Relax_drop_term_analysis_second_revision_without_advec_new.png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_9_2_moving_drop_term_analysis_smaller_drop_theta_5_better.png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_9_3_dewetting_term_analysis_no_legend_new_without_advec.png}
    \caption{Time evolution of the 
  $\mathbf{L}_2$ norm (as defined by equation 
  (\ref{eq:magnitude})) of the $x$-component of the terms appearing on the right hand side of the second of equations (\ref{eq:hydro2}). The plot is in log-lin scale.
  The panels refer to three different numerical experiments: (a) spreading droplet, (b) sliding droplet and (c) thin film dewetting. The symbols correspond to: 
  film pressure gradient, $-h\partial_x p_{\mbox{\tiny{film}}}$, (\textcolor{pyred}{$\star$}); friction, $-\nu \alpha(h) u_x$, (\textcolor{pyblue}{$\bullet$}); longitudinal dissipation terms 
  $\nu \nabla^2 (h u_x) + 2\nu \partial_x (\nabla \cdot (h\mathbf{u}))$, (\textcolor{pyorange}{$\blacktriangle$}).}
  \label{fig:Referee_1}
\end{figure}
